---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Update of RELEASE-2026

Note that Fish and By catch herp have not been updated this time. By catch herps likely will be removed in future release given their design was not suitable for diversity data. Fish data are currently under updating by NEON.

# neonDivData

<!-- badges: start -->
<!-- badges: end -->

The goal of neonDivData is to provide cleaned NEON organismal data to facilitate biodiversity research. The authors of this R data package have all spend lots of effort to clean NEON data for our own research; it makes the most sense to document such processes and provide the clean data product so that the large community can use them readily. This will save us time to dig into the extensive documenations of NEON data and to clean up the data individually. 

Data products for all taxonomic groups are long data frames with names in the format of `data_xxx` (e.g. `data_plant`, `data_fish`). Location inforamtions are in `neon_location`; taxonomic informations are in `neon_taxa`.

Despite the aim of this R data package is to facilitate biodiversity research, we keep `NA`s in data products so that information about sampling effort is saved. For some taxonomic groups, we removed observations with above genus level species identification. All codes to clean up the NEON data are available within our [Github repo](https://github.com/daijiang/neonDivData/tree/master/data-raw). Users can use them to customize their own data product if needed.

Available taxonomic groups and their brief summaries:

```{r}
# sites too long, don't print
knitr::kable(neonDivData::data_summary[, !names(neonDivData::data_summary) %in% c("sites", "data_package_id")])
```

## Installation

You can install the development version of neonDivData from [Github](https://github.com/daijiang/neonDivData) with:

``` r
install.packages("neonDivData", repos = 'https://daijiang.r-universe.dev')
```

## Available data products

```{r example}
neonDivData::neon_sites
neonDivData::neon_taxa
neonDivData::neon_location
neonDivData::data_algae
neonDivData::data_beetle
neonDivData::data_bird
neonDivData::data_macroinvertebrate
neonDivData::data_mosquito
neonDivData::data_plant
neonDivData::data_small_mammal
neonDivData::data_tick
neonDivData::data_tick_pathogen
```

# Annual data update

NEON releases new data each January. The package is updated annually, typically in March, by running the pipeline in `data-raw/` locally and pushing the rebuilt package data to GitHub.

**Prerequisites:** a NEON API token (register at [data.neonscience.org](https://data.neonscience.org)) stored in your `.Renviron` as `NEON_TOKEN=your_token_here`.

**Steps:**

```bash
# 1. Remove the download sentinel to trigger a fresh download
rm -f data-raw/NEON_raw_data/.last_download

# 2. Run the full pipeline (download ~1-2 days, cleaning ~1 hour)
make all

# 3. Rebuild documentation and check the package
Rscript -e "devtools::document()"
make check
```

`make all` runs three scripts in sequence, skipping any step whose outputs are already up to date:

| Step | Script | What it does |
|------|--------|--------------|
| 1 | `data-raw/01_download_data.R` | Downloads all 10 NEON data products via `neonUtilities`; saves timestamped RDS files |
| 2 | `data-raw/02_clean_save_data.R` | Cleans each taxon and saves `data/*.rda` package objects |
| 3 | `data-raw/03_plant_names.R` | Standardizes plant names against WCVP; updates `data/data_plant.rda` and `data-raw/plant_name_lookup.csv` |

If the download is interrupted, re-run `make all` and it will resume from step 2 if the RDS files are present, or repeat only the failed taxon downloads if you re-run step 1 manually. To force a specific step without re-running earlier ones, use `make download`, `make clean-data`, or `make plants`.

# Contributing

Contributions are welcome. You can provide comments and feedback or ask questions by filing an issue on Github [here](https://github.com/daijiang/neonDivData/issues) or making pull requests.

# Code of conduct

Please note that the 'neonDivData' project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

