---
title: "Appendix"
author: Daijiang Li, Sydne Record, Eric Sokol, Matthew E. Bitters, Melissa Y. Chen,  
  Y. Anny Chung, Matthew R. Helmus, Ruvi Jaimes, Lara Jansen, Marta A. Jarzyna, Michael
  G. Just, Jalene M. LaMontagne, Brett Melbourne, Wynne Moss, Kari Norman, Stephanie
  Parker, Natalie Robinson, Bijan Seyednasrollah, Colin Smith, Sarah Spaulding, Thilina
  Surasinghe, Sarah Thomsen, Phoebe Zarnetske
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purposes of this appendix include:

- Print the header of each data product provided by this project
- Describe the meaning of each variable in these data products
- Provide the R codes used for the figures included in this study 

Such information are also available on the project website (https://daijiang.github.io/neonDivData/reference/index.html) and project GitHub repository (https://github.com/daijiang/neonDivData).

```{r}
library(neonDivData)
library(DT)
```

# Periphyton, seston, and phytoplankton collection

```{r}
DT::datatable(head(data_algae, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `sampleCondition`: Condition of samples.
- `perBottleSampleVolume`: Per bottle sample volumn (milliliter).
- `release`: Version of data release by NEON.
- `habitatType`: Habitat type sampled.
- `algalSampleType`: Type of algal sample collected.
- `benthicArea`: Area of the benthos sampled (square Meter).
- `samplingProtocolVersion`: The NEON document number and version where detailed information regarding the sampling method used is available; format NEON.DOC.######vX.
- `substratumSizeClass`: Size class of the substratum sampled.
- `samplerType`: Type of sampler used to collect the sample.
- `phytoDepth1`: First phytoplankton sample depth (meter) at sampling location
- `phytoDepth2`: Second phytoplankton sample depth (meter) at sampling location
- `phytoDepth3`: Third phytoplankton sample depth (meter) at sampling location
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.


# Ground beetles sampled from pitfall traps


```{r}
DT::datatable(head(data_beetle, n = 30), options = list(scrollX = T))
```


### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `trapID`: Identifier for trap.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `boutID`: Identifier for bout.
- `nativeStatusCode`: The process by which the taxon became established in the location.
'A': Presumed absent, due to lack of data indicating a taxon's presence in a given location;
'N': Native; 'I': Introduced; 'UNK': Status unknown.
- `release`: Version of data release by NEON.
- `remarks`: Remarks (techinical notes) of record.
- `samplingProtocolVersion`: The NEON document number and version where detailed information regarding the sampling method used is available; format 'NEON.DOC.######vX'.
- `trappingDays`: Decimal days between trap setting and collecting events.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.


# Breeding landbird point counts data

```{r}
DT::datatable(head(data_bird, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `pointID`: Identifier for a point location.
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`. `NA` represents no bird observed.
- `unit`: Unit of the values in the `value` column.
- `pointCountMinute`: The minute of sampling within the point count period.
- `targetTaxaPresent`: Indicator of whether the sample contained individuals of the target taxa.
- `nativeStatusCode`: The process by which the taxon became established in the location.
'A': Presumed absent, due to lack of data indicating a taxon's presence in a given location;
'N': Native; 'I': Introduced; 'UNK': Status unknown.
- `observerDistance`: Radial distance between the observer and the individual(s) being observed (unit: meter).
- `detectionMethod`: How the individual(s) was (were) first detected by the observer.
- `visualConfirmation`: Whether the individual(s) was (were) seen after the initial detection.
- `sexOrAge`: Sex of individual if detectable, age of individual if individual can not be sexed.
- `release`: Version of data release by NEON.
- `startCloudCoverPercentage`: Observer estimate of percent cloud cover at start of sampling.
- `endCloudCoverPercentage`: Observer estimate of percent cloud cover at end of sampling.
- `startRH`: Relative humidity as measured by handheld weather meter at the start of sampling.
- `endRH`: Relative humidity as measured by handheld weather meter at the end of sampling.
- `observedHabitat`: Observer assessment of dominant habitat at the sampling point at sampling time.
- `observedAirTemp`: The air temperature (celsius) measured with a handheld weather meter.
- `kmPerHourObservedWindSpeed`: The average wind speed measured with a handheld weather meter, in kilometers per hour.
- `samplingProtocolVersion`: The NEON document number and version where detailed information regarding the sampling method used is available; format 'NEON.DOC.######vX'.
- `remarks`: Remarks of record.
- `clusterCode`: Alphabetic code (A-Z) linked to clusters (groups of individuals of the same species) spanning multiple records. It is only used to link clusters that take up multiple lines on the data sheet.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.

# Fish survey data collected by NEON

```{r}
DT::datatable(head(data_fish, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `pointID`: NEON sampling point identifier.
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `reachID`: An identifier for the set of information associated with the reach.
- `samplerType`: Type of sampler used to collect the sample.
- `fixedRandomReach`:	An indication of whether the reach is fixed or random.
- `measuredReachLength`:	The length of the reach as measured by the technicians when the reach was established (meters).
- `efTime`: Operational time of the electrofisher (second).
- `passStartTime`: The start time of the pass.
- `passEndTime`:	The end time of the pass.
- `mean_efishtime`: Average efish time (in second).
- `release`:	Version of data release by NEON.
- `netSetTime`: Time the net was set.
- `netEndTime`: Time the net was pulled.
- `netDeploymentTime`:	Total time of deployment of the net (hours).
- `netLength`: Length of the net (meter).
- `netDepth`:	Deployment depth of the net (meter).
- `efTime2`:	Operational time of the electrofisher for the second electrofisher (second).
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.


# Vertebrate Herpetofauna Bycatch sampled from pitfall traps

```{r}
DT::datatable(head(data_herp_bycatch, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `trapID`: Identifier for trap.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `trappingDays`: Cleaned up decimal days between trap setting and collecting events
- `release`: Version of data release by NEON.
- `nativeStatusCode`: 	The process by which the taxon became established in the location only provided for vert bycatch herp
- `remarksSorting`: Technician notes; free text comments accompanying the record from sorting table
- `remarksFielddata`: Technician notes; free text comments accompanying the record from fielddata table
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.

# Macroinvertebrate data

```{r}
DT::datatable(head(data_macroinvertebrate, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `estimatedTotalCount`: Estimated total count.
- `individualCount`: Individual count.
- `subsamplePercent`: Percent of the total sample contained in the subsample.
- `release`: Version of data release by NEON.
- `benthicArea`: Area sampled (square meter).
- `habitatType`: Habitat type sampled.
- `samplerType`: Type of sampler used to collect the sample.
- `substratumSizeClass`: Size class of the substratum sampled.
- `remarks`: Remarks of record.
- `ponarDepth`: Depth (meter) of petite ponar sample.
- `snagLength`: Length (centimeter) of snag sampled.
- `snagDiameter`: Diameter (centimeter) of snag sampled.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.

# Mosquitoes sampled from CO2 traps

```{r}
DT::datatable(head(data_mosquito, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `sampleID`: Identifier for sample.
- `subsampleID`: Unique identifier associated with each subsample per sampleID.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `nativeStatusCode`: The process by which the taxon became established in the location.
'A': Presumed absent, due to lack of data indicating a taxon's presence in a given location;
'N': Native; 'I': Introduced; 'UNK': Status unknown.
- `release`: Version of data release by NEON.
- `samplingProtocolVersion`: The NEON document number and version where detailed information regarding the sampling method used is available; format 'NEON.DOC.######vX'.
- `sex`: M for male, F for female, U for unknown.
- `sortDate`: Date sample was sorted.
- `subsampleWeight`:	Weight of subsample in gram.
- `totalWeight`: Weight of entire sample in gram.
- `trapHours`: Number of hours between trap setting and collecting events.
- `weightBelowDetection`: Notes regarding the weight relative to scale detection limit.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.

# Plant survey data collected by NEON

```{r}
DT::datatable(head(data_plant, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `subplotID`: This is the NEON provided subplot ID in the format of `subplot_id`, then `subsubplot_id` and 1 or 10 (m^2); if the sampling unit is 100 m^2, the values are 31, 32, 40, and 41.
- `subplot_id`: Subplot ID; each plot normally has four 100 m^2 subplots (31, 32, 40, 41).
- `subsubplot_id`: Subsubplot ID (1, 2, 3, 4) for sampling units at 1 m^2 or 10 m^2.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen. Values are 'genus', 'species', 'speciesGroup', 'subSpecies', or 'variety.' Species accounts for the majority of the entries. _Higher ranks have already been filtered out_ because we think they are too vague for biodiversity research.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`. If the individual was observed out of 1 square meter subplots, the value will be `NA` (i.e., _presence only_).
- `unit`: Unit of the values in the `value` column.
- `presence_absence`: All 1s since every record represent a species.
- `boutNumber`: Number of sampling bout, most sites sample only 1 bout.
- `nativeStatusCode`: Whether the species is a native or non-native species. 'A': Presumed absent, due to lack of data indicating a taxon's presence in a given location; 'N': Native; 'N?': Probably Native; 'I': Introduced; 'I?': Probably Introduced; 'NI': Native and Introduced, some infrataxa are native and others are introduced; 'NI?': Probably Native and Introduced, some infrataxa are native and others are introduced; 'UNK': Status unknown.
- `heightPlantOver300cm`: Indicator of whether individuals of the species in the sample are taller than 300 cm.
- `heightPlantSpecies`: Ocular estimate of the height (centimeter) of the plant species (if height is < 300 cm).
- `release`: Version of data release by NEON.
- `sample_area_m2`: The area of the sampling unit that the observed plant was located in. Potential values are 1, 10, or 100.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.

# Small mammal box trap data collected by NEON

```{r}
DT::datatable(head(data_small_mammal, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `year`: Observation year.
- `month`: Observation month.
- `n_trap_nights_per_bout_per_plot`: Number of trap nights per bout per plot.
- `n_nights_per_bout`: Number of nights per bout.
- `nativeStatusCode`: Whether the species is a native or non-native species.
'A': Presumed absent, due to lack of data indicating a taxon's presence in a
given location; 'N': Native; 'I': Introduced; 'UNK': Status unknown.
- `release`: Version of data release by NEON.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.

# Ticks sampled using drag cloths

```{r}
DT::datatable(head(data_tick, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `LifeStage`: Life stage of the sample ('Adult', 'Larva', or 'Nymph').
- `release`: Version of data release by NEON.
- `remarks_field`: Technician notes; free text comments accompanying the record.
- `samplingMethod`: Name or code for the method used to collect or test a sample.
- `targetTaxaPresent`: Indicator of whether the sample contained individuals of the target taxa ('Y' or 'N').
- `totalSampledArea`: Total area sampled (square Meter).
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.

# Tick-borne pathogen status

```{r}
DT::datatable(head(data_tick_pathogen, n = 30), options = list(scrollX = T))
```

### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `plotID`: Plot identifier (NEON site code_XXX).
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `lifeStage`: Life stage of the host (all Nymph).
- `testProtocolVersion`: The protocol version used to test the sample.
- `release`: Version of data release by NEON.
- `n_tests`: Number of tests conducted.
- `n_positive_test`: Number of tests that were positive.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.
- `nlcdClass`: National Land Cover Database Vegetation Type Name.
- `plotType`: NEON plot type in which sampling occurred: tower, distributed or gradient.

# Zooplankton density data

```{r}
DT::datatable(head(data_zooplankton, n = 30), options = list(scrollX = T))
```


### Definitions of columns

- `location_id`: Location id.
- `siteID`: NEON site code.
- `unique_sample_id`: Identity of unique samples, usually it has location and date information.
- `observation_datetime`: Observation date and time.
- `taxon_id`: Accepted species code, based on one or more sources.
- `taxon_name`:	Scientific name, associated with the taxonID. This is the name
 of the lowest level taxonomic rank that can be determined.
- `taxon_rank`: The lowest level taxonomic rank that can be determined for the individual or specimen.
- `variable_name`: The variable name(s) represented by the `value` column.
- `value`: Value of the variable(s) specified by `variable_name`.
- `unit`: Unit of the values in the `value` column.
- `release`: Version of data release by NEON.
- `samplerType`:	Type of sampler used to collect the sample.
- `towsTrapsVolume`: Sample volume (liter) collected for zooplankton.
- `latitude`: The geographic latitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `longitude`: The geographic longitude (in decimal degrees, WGS84) of the geographic center of the reference area.
- `elevation`: Elevation (in meters) above sea level.

# R code to generate figures in the manuscript

```{r, code = readLines("figures.R"), eval = FALSE}

```

